\bf Approximating Likelihood Ratios with Calibrated Discriminative Classifiers

### 2016-01-22 jasa/paper-jasa.tex
Section 3.3: added some more inline comments

As a concrete example, let us consider searches for new particles at the Large
Hadron Collider (LHC). The simulator that is sampling from $p(\mathbf{x}|\theta)$ is
based on quantum field theory, a detailed simulation of the particle detector,
and data processing algorithms that transform raw sensor data into the feature
vector $\mathbf{x}$~\citep{Sjostrand:2006za,Agostinelli:2002hh}. The ATLAS and CMS
experiments have published  hundreds papers where the final result was
formulated as a hypothesis test or confidence interval using a generalized
likelihood ratio test~\citep{Cowan:2010js}, including most notably the discovery
of the Higgs boson~\citep{Aad:2012tfa,Chatrchyan:2012ufa} and subsequent
measurement of its properties. The bulk of the likelihood ratio tests at the LHC
are based on the distribution of a single event-level feature that discriminates
between a hypothesized process of interest (labeled \textit{signal}) and various
other processes (labeled \textit{background}). Typically, pseudo-data from the
simulator are used to approximate the density at various parameter points, and
an interpolation algorithm is used to approximate the parametrized
model~\citep{Cranmer:2012sba}. In particular, to improve the statistical power
of these tests, hundreds of these searches have already been using supervised
learning to train discriminative classifiers that take advantage of a high
dimensional feature vector $\mathbf{x}$. \glnote{We should then elaborate on how the
proposed method could help these searches.}

@@

Let $\mathbf{X}$ be a random vector with values $\mathbf{x} \in {\cal X}
\subseteq \mathbb{R}^p$ and let $p_\mathbf{X}(\mathbf{x}|\theta)$ denote the
density probability of $\mathbf{X}$ at value $\mathbf{x}$ under the
parametrization $\theta$. Let also assume some i.i.d. observed data ${\cal D} = \{
\mathbf{x}_1, \dots, \mathbf{x}_n \}$. In the setting where one is interested in
simple hypothesis testing between a null $\theta=\theta_0$ against an alternate
$\theta=\theta_1$, the Neyman-Pearson lemma states that the likelihood ratio
\begin{equation}\label{eqn:likelihood-ratio-test}
T({\cal D}; \theta_0, \theta_1) = \prod_{\mathbf{x} \in {\cal D}} \frac{ p_\mathbf{X}(\mathbf{x}|\theta_0)}{ p_\mathbf{X}(\mathbf{x}|\theta_1)}
\end{equation}
is the most powerful test statistic.

@@

The main result of this paper (see Theorem~\ref{thm:ratio-equivalence}) is to
generalize the observation that one can form a test statistic
\begin{equation}\label{eqn:likelihood-ratio-test-equiv}
T'({\cal D}; \theta_0, \theta_1) = \prod_{\mathbf{x} \in {\cal D}} \frac{ p_\mathbf{Y}(y=s(\mathbf{x}) | \theta_0)}{ p_\mathbf{Y}(y=s(\mathbf{x}) | \theta_1)}
\end{equation}
that is strictly equivalent to \ref{eqn:likelihood-ratio-test}, provided the change
of variable $\mathbf{Y} = s(\mathbf{X})$ is based
on a (parametrized) function $s$ that is strictly monotonic with the density ratio
\begin{equation}
r(\mathbf{x};\theta_0, \theta_1) = \frac{p_\mathbf{X}(\mathbf{x}|\theta_0)}{p_\mathbf{X}(\mathbf{x}|\theta_1)}.
\end{equation}
As derived below, this allows to recast the original likelihood ratio test into
an alternate form in which supervised learning can be used to build
$s(\mathbf{x})$ as a discriminative classifier.  In
Section~\ref{sec:generalized-likelihood-ratio} we extend this result to
generalized likelihood ratio tests, where it will be useful to have the
classifier  parametrized in terms of $(\theta_0, \theta_1)$.

@@

\begin{theorem}
    \label{thm:ratio-equivalence}
    Let $\mathbf{X}$ be a random vector vector with values in ${\cal X} \subseteq \mathbb{R}^p$ and parametrized probability
    density $p_{\mathbf{X}}(\mathbf{x} = (x_1, ..., x_p)|\theta)$ and let
    $s : \mathbb{R}^p \mapsto \mathbb{R}$ be a function monotonic with the density ratio
    $r(\mathbf{x};\theta_0,\theta_1) = \frac{p_\mathbf{X}(\mathbf{x}|\theta_0)}{p_\mathbf{X}(\mathbf{x}|\theta_1)}$,
    for given parameters $\theta_0$ and $\theta_1$. In these conditions,
    \begin{equation}
        r(\mathbf{x};\theta_0,\theta_1) = \frac{p_\mathbf{X}(\mathbf{x}|\theta_0)}{p_\mathbf{X}(\mathbf{x}|\theta_1)} = \frac{p_\mathbf{Y}(y=s(\mathbf{x})|\theta_0)}{p_\mathbf{Y}(y=s(\mathbf{x})|\theta_1)},
    \end{equation}
    where $p_\mathbf{Y}(y=s(\mathbf{x};\theta_0,\theta_1)|\theta)$ is the induced probability density of
    $\mathbf{Y} = s(\mathbf{X};\theta_0,\theta_1)$.
\end{theorem}

@@

\subsection{Probabilistic classification for approximating ratios}
\label{sec:clf-for-ratios}

@@

For $P(Y=0)=P(Y=1)=\frac{1}{2}$, the best regression function $s^*$ simplifies
to
\begin{equation}
s^*(\mathbf{x}) = \frac{p_{\mathbf{X}|Y}(\mathbf{x}|Y=1)}{p_{\mathbf{X}|Y}(\mathbf{x} | Y=0) + p_{\mathbf{X}|Y}(\mathbf{x} | Y=1)}.
\end{equation}
If we further assume that samples for $Y=0$ are drawn from some parametrized
distribution with probability density $p_{\mathbf{X}}(\mathbf{x}|\theta_0)$
and similarly for $Y=1$, then the best regression function can be rewritten
as
\begin{equation}
s^*(\mathbf{x}) = \frac{p_{\mathbf{X}}(\mathbf{x}|\theta_1)}{p_{\mathbf{X}}(\mathbf{x} | \theta_0) + p_{\mathbf{X}}(\mathbf{x} | \theta_1)}.
\end{equation}
In particular, this regression function satisfies conditions of
Theorem~\ref{thm:ratio-equivalence} since $s^*(\mathbf{x}) =
m(\frac{p_\mathbf{X}(\mathbf{x}|\theta_0)}{p_\mathbf{X}(\mathbf{x}|\theta_1)})$,
for $m(r(\mathbf{x})) = \frac{1}{1 + r(\mathbf{x})}$, is monotonic with
$\frac{p_\mathbf{X}(\mathbf{x}|\theta_0)}{p_\mathbf{X}(\mathbf{x}|\theta_1)}$.
In other words, Proposition~\ref{thm:best-classifier} yields a sufficient
procedure for Theorem~\ref{thm:ratio-equivalence} to hold, guaranteeing that any
{\it universally strongly consistent} algorithm can be used for learning $s^*$.
Note however, that it is not a necessary procedure since
Theorem~\ref{thm:ratio-equivalence} holds for any monotonic function $m$ of the
density ratio, i.e., not for $m(r(\mathbf{x})) = \frac{r(\mathbf{x})}{1 +
r(\mathbf{x})}$ only.

Equivalently, since $m^{-1}(s^*(\mathbf{x})) = \frac{1 -
s^*(\mathbf{x})}{s^*(\mathbf{x})} = r(\mathbf{x})$,
Proposition~\ref{thm:best-classifier} shows that the likelihood ratio estimation
problem is directly related to probabilistic classification, and that algorithms
proposed to solve one can be used for solving the other. In this context,
Theorem~\ref{thm:ratio-equivalence} shows that in case we learn a probabilistic
classifier $s(\mathbf{x})$ which is imperfect up to a monotonic transformation
of $r(\mathbf{x})$, then one can still resort to calibration (i.e., modelling
$p_Y(y=s(\mathbf{x}))$) to compute $r(\mathbf{x})$ exactly.

\subsection{Learning the correct mapping and its distribution}

\glnote{This section needs to be improved. Not sure we need a section on its own for that either.}

In order for this approach to be useful in the likelihood-free setting, where it
is not possible to evaluate the density $p(\mathbf{x}|\theta)$, we need to be
able to approximate both $s(\mathbf{x})$ and
$p(s(\mathbf{x})|\theta)$ based on samples $\{x_i\}$ drawn from the generative
model $p(\mathbf{x}|\theta)$.  Denote the dimensionality reduction map
$\hat{s}(x)$ and its distribution $\hat{p}(\hat{s}|\theta)$.

One strength of this approach is that it factorizes the approximation of the
per-event likelihood ratio ($\hat{s}(\mathbf{x}) \approx
s(\mathbf{x})$) from the calibration procedure ($\hat p(\hat
s| \theta) \approx p(\hat{s}|\theta)$). Thus, even if the classifier does a poor
job of reproducing the level sets of the per-event likelihood ratio, the density
of $\hat{s}$ can still be well calibrated. In that case, one might loose power,
but the resulting inference will still be valid. This point was made by
\cite{Neal:2007zz} and is well appreciated by the particle physics community
that typically takes a conservative attitude towards the use of machine learning
classifiers precisely due to concerns about the calibration $p$-values in the
face of nuisance parameters associated to the simulator.

@@

Evaluating the generalized likelihood ratio as defined by
Eqn.~\ref{eqn:generalized-lr} requires finding for both the numerator and the
denominator the maximum likelihood estimator
\begin{equation}\label{eq:mle}
    \hat{\theta} = \argmax_\theta p({\cal D} | \theta).
\end{equation}
Again, this is made difficult in the likelihood-free setting and it is not
obvious that we can find the same estimators if we are working instead with
$p(s(\mathbf{x})|\theta)$. Fortunately, there is a construction
based on $s$ that works: the maximum likelihood estimate of Eqn.~\ref{eq:mle} is
the same as the value that maximizes the likelihood ratio with respect to
$p({\cal D}|\theta_1)$, for some fixed value of $\theta_1$. This allows us to
use Theorem~\ref{thm:ratio-equivalence} to reformulate the maximum likelihood
estimate as
\begin{align}\label{eq:mle_withs}
\hat{\theta} &= \argmax_\theta  p({\cal D} | \theta) \nonumber \\
&= \argmax_\theta  \prod_{\mathbf{x} \in {\cal D}} \frac{p(\mathbf{x}| \theta)}{p(\mathbf{x}|\theta_1)} \nonumber \\
&= \argmax_\theta  \prod_{\mathbf{x} \in {\cal D}} \frac{p(s(\mathbf{x}; \theta, \theta_1) | \theta)}{p(s(\mathbf{x}; \theta, \theta_1) |\theta_1)} \;,
\end{align}
where $s(\mathbf{x};\theta,\theta_1)$ denotes a \textit{parametrized}
transformation $s$ of $\mathbf{X}$ in terms of $(\theta,\theta_1)$ that is monotonic
with $r(\mathbf{x};\theta,\theta_1)$. Note that it is important that we include
the denominator $p(s(\mathbf{x}; \theta, \theta_1) |\theta_1)$ because this
cancels Jacobian factors that vary with $\theta$.

\subsection{Parametrized classifier}

In order to provide parameter inference in the likelihood-free setting as
described above, we must train a family $s(\mathbf{x};\theta,\theta_1)$ of classifiers parametrized by
$\theta_0$ and $\theta_1$, the parameters associated to the null and alternate
hypotheses, respectively. While this could be done independently for all
$\theta_0$ and $\theta_1$, using the procedure outlined in
Section~\ref{sec:clf-for-ratios}, it is desirable and convenient to have a
smooth evolution of the classification score as a function of the parameters.
For this reason, we anticipate a single learning stage based on training data with input
$(\mathbf{x}, \theta_0, \theta_1)_i$ and target $y_i$, as outlined in Algorithm~\ref{alg:training}. Somewhat unusually, the
unknown values of the parameters are taken as input to the classifier; their
values will be specified via the enveloping (generalized) likelihood ratio of
Eqn.~\ref{eqn:generalized-lr}.

\begin{algorithm}[t]
\caption{Training of a parametrized classifier.}\label{alg:training}
    \begin{algorithmic}
        \State ${\cal L := \{ \}}$
        \For{ $\theta_0$ in $\Theta_0$ }
        	\For{ $\theta_1$ in $\Theta_1$ }
        		\State generate $\mathbf{x} \sim p(\mathbf{x}|\theta_0)$
        		\State append $\{ (\mathbf{x}, \theta_0, \theta_1, y=0) \}$ to ${\cal L}$
        		\State generate $\mathbf{x} \sim p(\mathbf{x}|\theta_1)$
        		\State append $\{ (\mathbf{x}, \theta_0, \theta_1, y=1) \}$ to ${\cal L}$
        	\EndFor
        \EndFor
        \State Learn $s(\mathbf{x}; \theta_0, \theta_1)$ from ${\cal L}$
    \end{algorithmic}
\end{algorithm}

@@

\subsection{Parametrized calibration}

Once the classifier is trained, we can use the generative model together with a
univariate density estimation technique (e.g. histograms or kernel density
estimation) to approximate $p(s|\theta)$ for specific parameter
points. For a single parameter point, this is a tractable univariate density
estimation problem. The challenge comes from the need to calibrate this density
for all values of $\theta$. A straight forward approach would be to run the
generative model on demand for any particular value of $\theta$. In the context
of a likelihood fit this would mean that the optimization algorithm that is
trying to maximize the likelihood with respect to $\theta$ needs access to the
generative model $p(\mathbf{x}|\theta)$. This can be  impractical when the generative
model is computationally expensive or has high-latency (for instance some human
intervention is required to reconfigure the generative model).  In HEP, with a
fixed classifier, it has become common  to interpolate the distribution between
discrete values of $\theta$ in order to produce a continuous parametrization for
$p(s | \theta)$~\citep{Cranmer:2012sba}.
One can easily imagine a number of approaches to embedding the classifier and
estimating the density $p(s|\theta)$ and the relative merits of those
approaches will depend critically on the dimensionality of $\theta$ and the
computational cost of the generative model. We leave a more general strategy for
this overarching optimization problem as an area of future work.

@@

 Once the classifier is trained, large samples of pseudo-data are drawn from
 $p_s(x | \nu)$ and $p_b(x | \nu)$ and we estimate the distributions
 $\hat{p}_s(\hat s | \nu)$ and $\hat{p}_b(\hat s | \nu)$ continuously
 parametrized in $\nu$. An example of the distributions of $\hat s$ for the
 signal and background events with $\nu=\nu_0$ is shown in
 Figure~\ref{fig:tmva}.

@@

Nuisance parameters are an after thought in the typical usage of machine
learning in HEP. In fact, most discussions would related to the training and
optimizing the classifier only consider $p_b(x)$ and $p_s(x)$ with $\nu=\nu_0$
being implicit. However, as experimentalists we know that we must account for
various forms of systematic uncertainty, parametrized by nuisance parameters
$\nu$. In practice, we take the classifier as fixed and then propagate
uncertainty through the classifier as in Eq.~\ref{eq:typicalML}. Building the
distribution $p(\hat s|\nu)$ for values of $\nu$ other than the nominal $\nu_0$
used to train the classifier can be thought of as a calibration necessary for
classical statistical inference; however, this classifier is clearly not optimal
for $\nu \ne \nu_0$.

@@

The standard use of machine learning in HEP can be improved by training a
parametrized classifier corresponding to the generalized likelihood ratio test
\begin{equation}
\lambda(\mu) = \frac{p(D|\mu, \hat{\hat{\nu}})}{p(D|\hat \mu, {\hat{\nu}})} \;,
\end{equation}
following the approach outlined in Section~\ref{S:GLR}.

There is an interesting distinction between this approach and the standard use
in which the classifier is trained for a fixed $\nu_0$. In the standard use one
trains a classifier for signal vs. background, which is equivalent (in an ideal
setting) to training a classifier for  null (background-only) vs. alternate
(signal-plus-backgound) since the resulting regression functions are one-to-one
with each other. In contrast, in the case of the generalized likelihood ratio
test
\begin{equation}\label{eq:hep_improved}
 \frac{p(x| 0, \hat{\hat{ \nu}})}{p(x|\hat \mu, \hat\nu)} =  \frac{p_b(x| \hat{\hat{ \nu}})}{ \hat \mu p_s( x_e \, |\,  \hat\nu)  + (1- \hat \mu )\, p_b( x_e \,|\, \hat \nu)} \; ,
\end{equation}
the background components don't cancel and there is an additional term $p_b(x|
\hat{\hat{ \nu}})/p_b(x| {\hat{ \nu}})$. In practice, with classifiers of finite
capacity, there will be some tradeoff between taking into account this
additional term and the more challenging learning problem when $\mu$ is very
small.

@@

While the original motivation for this work was to improve the treatment of
systematic uncertainties in new particle searches by parametrizing the
classifier in terms of the nuisance parameters $\nu$, the same approach can be
used for parameter inference. In the case of new particle searches the parameter
of interest is the mixture coefficient for the signal component $p_s(x|\nu)$.
When measuring particle properties the distribution of the features also depend
on parameters such as a particle's mass and quantum numbers. This is easily
accommodated by extending $p_s(x|\nu) \to p_s(x|\theta)$, where $\theta$
includes both parameters of interest and nuisance parameters.

This formalism represents a significant step forward in the usage of machine
learning in HEP, where classifiers have always been used between two static
classes of events and not parametrized explicitly in terms of the physical
quantities we wish to measure. The work of  ~\citep{Whiteson:2006ws} is similar
as the stochastic optimization was directly trying to minimize the measurement
uncertainty of a particle's mass; however, the resulting classifier was fixed.
This approach also offers the advantage that it explicitly reformulates the
per-experiment optimization to the per-event optimization, which is less
computationally intensive.

>>>>

As a concrete example, let us consider searches for new particles at the Large
Hadron Collider (LHC). The simulator that is sampling from $p(\mathbf{x}|\theta)$ is
based on quantum field theory, a detailed simulation of the particle detector,
and data processing algorithms that transform raw sensor data into the feature
vector $\mathbf{x}$~\citep{Sjostrand:2006za,Agostinelli:2002hh}. The ATLAS and CMS
experiments have published  hundreds papers where the final result was
formulated as a hypothesis test or confidence interval using a generalized
likelihood ratio test~\citep{Cowan:2010js}, including most notably the discovery
of the Higgs boson~\citep{Aad:2012tfa,Chatrchyan:2012ufa} and subsequent
measurement of its properties. The bulk of the likelihood ratio tests at the LHC
are based on the distribution of a single event-level feature that discriminates
between a hypothesized process of interest (labeled \textit{signal}) and various
other processes (labeled \textit{background}). Typically, pseudo-data from the
simulator are used to approximate the density at various parameter points, and
an interpolation algorithm is used to approximate the parameterized
model~\citep{Cranmer:2012sba}. In particular, to improve the statistical power
of these tests, hundreds of these searches have already been using supervised
learning to train discriminative classifiers that take advantage of a high
dimensional feature vector $\mathbf{x}$. \glnote{We should then elaborate on how the
proposed method could help these searches.}

@@

Let $\mathbf{X}$ be a random vector with values $\mathbf{x} \in {\cal X}
\subseteq \mathbb{R}^p$ and let $p_\mathbf{X}(\mathbf{x}|\theta)$ denote the
density probability of $\mathbf{X}$ at value $\mathbf{x}$ under the
parameterization $\theta$. Let also assume some i.i.d. observed data ${\cal D} = \{
\mathbf{x}_1, \dots, \mathbf{x}_n \}$. In the setting where one is interested in
simple hypothesis testing between a null $\theta=\theta_0$ against an alternate
$\theta=\theta_1$, the Neyman-Pearson lemma states that the likelihood ratio
\begin{equation}\label{eqn:likelihood-ratio-test}
T({\cal D}; \theta_0, \theta_1) = \prod_{\mathbf{x} \in {\cal D}} \frac{ p_\mathbf{X}(\mathbf{x}|\theta_0)}{ p_\mathbf{X}(\mathbf{x}|\theta_1)}
\end{equation}
is the most powerful test statistic.

@@

The main result of this paper (see Theorem~\ref{thm:ratio-equivalence}) is to
generalize the observation that one can form a test statistic
\begin{equation}\label{eqn:likelihood-ratio-test-equiv}
T'({\cal D}; \theta_0, \theta_1) = \prod_{\mathbf{x} \in {\cal D}} \frac{ p_\mathbf{Y}(y=s(\mathbf{x}) | \theta_0)}{ p_\mathbf{Y}(y=s(\mathbf{x}) | \theta_1)}
\end{equation}
that is strictly equivalent to \ref{eqn:likelihood-ratio-test}, provided the change
of variable $\mathbf{Y} = s(\mathbf{X})$ is based
on a (parameterized) function $s$ that is strictly monotonic with the density ratio
\begin{equation}
r(\mathbf{x};\theta_0, \theta_1) = \frac{p_\mathbf{X}(\mathbf{x}|\theta_0)}{p_\mathbf{X}(\mathbf{x}|\theta_1)}.
\end{equation}
As derived below, this allows to recast the original likelihood ratio test into
an alternate form in which supervised learning can be used to build
$s(\mathbf{x})$ as a discriminative classifier.  In
Section~\ref{sec:generalized-likelihood-ratio} we extend this result to
generalized likelihood ratio tests, where it will be useful to have the
classifier  parameterized in terms of $(\theta_0, \theta_1)$.

@@

\begin{theorem}
    \label{thm:ratio-equivalence}
    Let $\mathbf{X}$ be a random vector vector with values in ${\cal X} \subseteq \mathbb{R}^p$ and parameterized probability
    density $p_{\mathbf{X}}(\mathbf{x} = (x_1, ..., x_p)|\theta)$ and let
    $s : \mathbb{R}^p \mapsto \mathbb{R}$ be a function monotonic with the density ratio
    $r(\mathbf{x};\theta_0,\theta_1) = \frac{p_\mathbf{X}(\mathbf{x}|\theta_0)}{p_\mathbf{X}(\mathbf{x}|\theta_1)}$,
    for given parameters $\theta_0$ and $\theta_1$. In these conditions,
    \begin{equation}
        r(\mathbf{x};\theta_0,\theta_1) = \frac{p_\mathbf{X}(\mathbf{x}|\theta_0)}{p_\mathbf{X}(\mathbf{x}|\theta_1)} = \frac{p_\mathbf{Y}(y=s(\mathbf{x})|\theta_0)}{p_\mathbf{Y}(y=s(\mathbf{x})|\theta_1)},
    \end{equation}
    where $p_\mathbf{Y}(y=s(\mathbf{x};\theta_0,\theta_1)|\theta)$ is the induced probability density of
    $\mathbf{Y} = s(\mathbf{X};\theta_0,\theta_1)$.
\end{theorem}

@@

\subsection{Probabilistic classification for likelihood ratios}
\label{sec:clf-for-ratios}

@@

For $P(Y=0)=P(Y=1)=\frac{1}{2}$, the best regression function $s^*$ simplifies
to
\begin{equation}
s^*(\mathbf{x}) = \frac{p_{\mathbf{X}|Y}(\mathbf{x}|Y=1)}{p_{\mathbf{X}|Y}(\mathbf{x} | Y=0) + p_{\mathbf{X}|Y}(\mathbf{x} | Y=1)}.
\end{equation}
If we further assume that samples for $Y=0$ are drawn from some parameterized
distribution with probability density $p_{\mathbf{X}}(\mathbf{x}|\theta_0)$
and similarly for $Y=1$, then the best regression function can be rewritten
as
\begin{equation}
s^*(\mathbf{x}) = \frac{p_{\mathbf{X}}(\mathbf{x}|\theta_1)}{p_{\mathbf{X}}(\mathbf{x} | \theta_0) + p_{\mathbf{X}}(\mathbf{x} | \theta_1)}.
\end{equation}
In particular, this regression function satisfies conditions of
Theorem~\ref{thm:ratio-equivalence} since $s^*(\mathbf{x}) =
m(\frac{p_\mathbf{X}(\mathbf{x}|\theta_0)}{p_\mathbf{X}(\mathbf{x}|\theta_1)})$,
for $m(r(\mathbf{x})) = \frac{1}{1 + r(\mathbf{x})}$, is monotonic with
$\frac{p_\mathbf{X}(\mathbf{x}|\theta_0)}{p_\mathbf{X}(\mathbf{x}|\theta_1)}$.
In other words, Proposition~\ref{thm:best-classifier} yields a sufficient
procedure for Theorem~\ref{thm:ratio-equivalence} to hold, guaranteeing that any
{\it universally strongly consistent} algorithm can be used for learning $s^*$.
Note however, that it is not a necessary procedure since
Theorem~\ref{thm:ratio-equivalence} holds for any monotonic function $m$ of the
density ratio, i.e., not for $m(r(\mathbf{x})) = \frac{r(\mathbf{x})}{1 +
r(\mathbf{x})}$ only.

Equivalently, since $m^{-1}(s^*(\mathbf{x})) = \frac{1 -
s^*(\mathbf{x})}{s^*(\mathbf{x})} = r(\mathbf{x})$,
Proposition~\ref{thm:best-classifier} shows that the likelihood ratio estimation
problem is directly related to probabilistic classification, and that algorithms
proposed to solve one can be used for solving the other. In this context,
Theorem~\ref{thm:ratio-equivalence} shows that in case we learn a probabilistic
classifier $s(\mathbf{x})$ which is imperfect up to a monotonic transformation
of $r(\mathbf{x})$, then one can still resort to calibration (i.e., modeling
$p_Y(y=s(\mathbf{x}))$) to compute $r(\mathbf{x})$ exactly.

\subsection{Approximating the reduction map and its distribution}

In order for this approach to be useful in the likelihood-free setting, we need
to be able to approximate both $s(\mathbf{x})$ and $p(s(\mathbf{x})|\theta)$
based on a finite number of samples $\{\mathbf{x}_i\}$ drawn from the generative
model $p(\mathbf{x}|\theta)$.
Let denote the approximated dimensionality
reduction map $\hat{s}(\mathbf{x})$ and its approximated distribution
$\hat{p}(\hat{s}(\mathbf{x})|\theta)$. \glnote{Explain how to approximate $p(s)$ in practice?}

One strength of this approach is that it factorizes the approximation of the
per-event likelihood ratio ($\hat{s}(\mathbf{x}) \approx s(\mathbf{x})$) from the
calibration procedure ($\hat p(\hat s(\mathbf{x})| \theta) \approx p(\hat{s}(\mathbf{x})|\theta)$).
Thus, even if the classifier does a poor job at reproducing the level sets of
the per-event likelihood ratio, the density of $\hat{s}$ can still be well
calibrated. In that case, one might loose power, but the resulting inference
will still be valid.\glnote{Shouldn't we elaborate?} This point was made by
\cite{Neal:2007zz} and is well appreciated by the particle physics community
that typically takes a conservative attitude towards the use of machine learning
classifiers precisely due to concerns about the calibration $p$-values in the
face of nuisance parameters associated to the simulator.

\glnote{We should comment as to what happens if $\hat{s}(x)$ is not monotonic with $r$}

\glnote{Add pseudo-code to summarize the whole pipeline?}

@@

Evaluating the generalized likelihood ratio as defined by
Eqn.~\ref{eqn:generalized-lr} requires finding for both the numerator and the
denominator the maximum likelihood estimator
\begin{equation}\label{eq:mle}
    \hat{\theta} = \argmax_\theta p({\cal D} | \theta).
\end{equation}
Again, this is made difficult in the likelihood-free setting and it is not
obvious that we can find the same estimators if we are working instead with
$p(s(\mathbf{x})|\theta)$. Fortunately, there is a construction
based on $s$ that works: the maximum likelihood estimate of Eqn.~\ref{eq:mle} is
the same as the value that maximizes the likelihood ratio with respect to
$p({\cal D}|\theta_1)$, for some fixed value of $\theta_1$. This allows us to
use Theorem~\ref{thm:ratio-equivalence} to reformulate the maximum likelihood
estimate as
\begin{align}\label{eq:mle_withs}
\hat{\theta} &= \argmax_\theta  p({\cal D} | \theta) \nonumber \\
&= \argmax_\theta  \prod_{\mathbf{x} \in {\cal D}} \frac{p(\mathbf{x}| \theta)}{p(\mathbf{x}|\theta_1)} \nonumber \\
&= \argmax_\theta  \prod_{\mathbf{x} \in {\cal D}} \frac{p(s(\mathbf{x}; \theta, \theta_1) | \theta)}{p(s(\mathbf{x}; \theta, \theta_1) |\theta_1)} \;,
\end{align}
where $s(\mathbf{x};\theta,\theta_1)$ denotes a \textit{parameterized}
transformation $s$ of $\mathbf{X}$ in terms of $(\theta,\theta_1)$ that is monotonic
with $r(\mathbf{x};\theta,\theta_1)$. Note that it is important that we include
the denominator $p(s(\mathbf{x}; \theta, \theta_1) |\theta_1)$ because this
cancels Jacobian factors that vary with $\theta$.

\subsection{Parameterized classifier}

In order to provide parameter inference in the likelihood-free setting as
described above, we must train a family $s(\mathbf{x};\theta,\theta_1)$ of classifiers parameterized by
$\theta_0$ and $\theta_1$, the parameters associated to the null and alternate
hypotheses, respectively. While this could be done independently for all
$\theta_0$ and $\theta_1$, using the procedure outlined in
Section~\ref{sec:clf-for-ratios}, it is desirable and convenient to have a
smooth evolution of the classification score as a function of the parameters.
For this reason, we anticipate a single learning stage based on training data with input
$(\mathbf{x}, \theta_0, \theta_1)_i$ and target $y_i$, as outlined in Algorithm~\ref{alg:training}. Somewhat unusually, the
unknown values of the parameters are taken as input to the classifier; their
values will be specified via the enveloping (generalized) likelihood ratio of
Eqn.~\ref{eqn:generalized-lr}.

\begin{algorithm}[t]
\caption{Training of a parameterized classifier.}\label{alg:training}
    \begin{algorithmic}
        \State ${\cal L := \{ \}}$
        \For{ $\theta_0$ in $\Theta_0$ }
        	\For{ $\theta_1$ in $\Theta_1$ }
        		\State generate $\mathbf{x} \sim p(\mathbf{x}|\theta_0)$
        		\State append $\{ (\mathbf{x}, \theta_0, \theta_1, y=0) \}$ to ${\cal L}$
        		\State generate $\mathbf{x} \sim p(\mathbf{x}|\theta_1)$
        		\State append $\{ (\mathbf{x}, \theta_0, \theta_1, y=1) \}$ to ${\cal L}$
        	\EndFor
        \EndFor
        \State Learn $s(\mathbf{x}; \theta_0, \theta_1)$ from ${\cal L}$
    \end{algorithmic}
\end{algorithm}

@@

\subsection{Parameterized calibration}

Once the classifier is trained, we can use the generative model together with a
univariate density estimation technique (e.g. histograms or kernel density
estimation) to approximate $p(s|\theta)$ for specific parameter
points. For a single parameter point, this is a tractable univariate density
estimation problem. The challenge comes from the need to calibrate this density
for all values of $\theta$. A straight forward approach would be to run the
generative model on demand for any particular value of $\theta$. In the context
of a likelihood fit this would mean that the optimization algorithm that is
trying to maximize the likelihood with respect to $\theta$ needs access to the
generative model $p(\mathbf{x}|\theta)$. This can be  impractical when the generative
model is computationally expensive or has high-latency (for instance some human
intervention is required to reconfigure the generative model).  In HEP, with a
fixed classifier, it has become common  to interpolate the distribution between
discrete values of $\theta$ in order to produce a continuous parameterization for
$p(s | \theta)$~\citep{Cranmer:2012sba}.
One can easily imagine a number of approaches to embedding the classifier and
estimating the density $p(s|\theta)$ and the relative merits of those
approaches will depend critically on the dimensionality of $\theta$ and the
computational cost of the generative model. We leave a more general strategy for
this overarching optimization problem as an area of future work.

@@

 Once the classifier is trained, large samples of pseudo-data are drawn from
 $p_s(x | \nu)$ and $p_b(x | \nu)$ and we estimate the distributions
 $\hat{p}_s(\hat s | \nu)$ and $\hat{p}_b(\hat s | \nu)$ continuously
 parameterized in $\nu$. An example of the distributions of $\hat s$ for the
 signal and background events with $\nu=\nu_0$ is shown in
 Figure~\ref{fig:tmva}.

@@

Nuisance parameters are an after thought in the typical usage of machine
learning in HEP. In fact, most discussions would related to the training and
optimizing the classifier only consider $p_b(x)$ and $p_s(x)$ with $\nu=\nu_0$
being implicit. However, as experimentalists we know that we must account for
various forms of systematic uncertainty, parameterized by nuisance parameters
$\nu$. In practice, we take the classifier as fixed and then propagate
uncertainty through the classifier as in Eq.~\ref{eq:typicalML}. Building the
distribution $p(\hat s|\nu)$ for values of $\nu$ other than the nominal $\nu_0$
used to train the classifier can be thought of as a calibration necessary for
classical statistical inference; however, this classifier is clearly not optimal
for $\nu \ne \nu_0$.

@@

The standard use of machine learning in HEP can be improved by training a
parameterized classifier corresponding to the generalized likelihood ratio test
\begin{equation}
\lambda(\mu) = \frac{p(D|\mu, \hat{\hat{\nu}})}{p(D|\hat \mu, {\hat{\nu}})} \;,
\end{equation}
following the approach outlined in Section~\ref{S:GLR}.

There is an interesting distinction between this approach and the standard use
in which the classifier is trained for a fixed $\nu_0$. In the standard use one
trains a classifier for signal vs. background, which is equivalent (in an ideal
setting) to training a classifier for  null (background-only) vs. alternate
(signal-plus-background) since the resulting regression functions are one-to-one
with each other. In contrast, in the case of the generalized likelihood ratio
test
\begin{equation}\label{eq:hep_improved}
 \frac{p(x| 0, \hat{\hat{ \nu}})}{p(x|\hat \mu, \hat\nu)} =  \frac{p_b(x| \hat{\hat{ \nu}})}{ \hat \mu p_s( x_e \, |\,  \hat\nu)  + (1- \hat \mu )\, p_b( x_e \,|\, \hat \nu)} \; ,
\end{equation}
the background components don't cancel and there is an additional term $p_b(x|
\hat{\hat{ \nu}})/p_b(x| {\hat{ \nu}})$. In practice, with classifiers of finite
capacity, there will be some trade-off between taking into account this
additional term and the more challenging learning problem when $\mu$ is very
small.

@@

While the original motivation for this work was to improve the treatment of
systematic uncertainties in new particle searches by parameterizing the
classifier in terms of the nuisance parameters $\nu$, the same approach can be
used for parameter inference. In the case of new particle searches the parameter
of interest is the mixture coefficient for the signal component $p_s(x|\nu)$.
When measuring particle properties the distribution of the features also depend
on parameters such as a particle's mass and quantum numbers. This is easily
accommodated by extending $p_s(x|\nu) \to p_s(x|\theta)$, where $\theta$
includes both parameters of interest and nuisance parameters.

This formalism represents a significant step forward in the usage of machine
learning in HEP, where classifiers have always been used between two static
classes of events and not parameterized explicitly in terms of the physical
quantities we wish to measure. The work of  ~\citep{Whiteson:2006ws} is similar
as the stochastic optimization was directly trying to minimize the measurement
uncertainty of a particle's mass; however, the resulting classifier was fixed.
This approach also offers the advantage that it explicitly reformulates the
per-experiment optimization to the per-event optimization, which is less
computationally intensive.
